---
title: "Class 3 Example: Exploring the Day 1 Survey"
author: "Thomas E. Love"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    code_folding: show
---

```{r setup, message=FALSE}
knitr::opts_chunk$set(comment=NA)
options(width = 70)
```

# Load the R Packages we need

```{r load_packages, message = FALSE, warning = FALSE}
library(magrittr); library(tidyverse)
## always need tidyverse, can include other packages too
```

# Load the Data

We will read in the .csv file of the data, and then pipe that result into the `tbl_df` function, which turns the data frame into a nicely organized *tibble*. 

- Since we've carefully stored the data file in the same directory as our R Project, we can read it in directly.

```{r load_data}
day1 <- read.csv("surveyday1_2018.csv") %>% tbl_df

day1
```

# Variables we're studying today

```{r} 
day1 %>%
    select(sex, favcolor, haircut, ageguess, pulse, lastsleep) %>%
    summary()
```

So we have missing values in the `favcolor`, `haircut`, `ageguess`, `pulse` and `lastsleep` data to watch out for.

# What are the favorite colors of students in 431?

We can get a count of the complete set of responses...

```{r}
day1 %>%
    count(favcolor)
```

We could restrict our interest just to the responses in 2018, and arrange them in descending order of the count.

```{r}
day1 %>%
	  filter(year == 2018) %>%
    count(favcolor) %>%
	  arrange(desc(n))
```

Since the data are categorical, we might use a *table* instead.

```{r}
day1 %>%
    select(favcolor) %>%
    table()
```

We could also do this with

```{r}
table(day1$favcolor)
```

It might be nice to see that table just for the 2018 data.

```{r}
day1 %>%
    filter(year == 2018) %>%
    select(favcolor) %>%
    table()
```

Let's drop the levels with zeros, and add a total.

```{r}
day1 %>%
    filter(year == 2018) %>%
    select(favcolor) %>%
    droplevels() %>%
    table() %>%
    addmargins()
```

## A more challenging question

What is the percentage of students who prefer `blue` in each year?

```{r}
day1 %>%
    group_by(year) %>%
    summarize(blue_count = sum(favcolor == "blue", na.rm=TRUE), 
              total_count = sum(!is.na(favcolor)), 
              blue_percent = round(100 * blue_count / total_count, 1)) 
```

What if we include some additional colors that are in the neighborhood of blue?

```{r}
day1 %>%
    group_by(year) %>%
    summarize(total_count = sum(!is.na(favcolor)), 
    		blue_count = sum(favcolor == "blue", na.rm=TRUE), 
    		blue_percent = round(100 * blue_count / total_count, 1),
    		blueplus_count = sum(favcolor %in% c("blue", "aqua", 
    						      "aquamarine", "light blue", "navy", 
    						      "royal blue", "sapphire", "sky blue", 
    						      "turquoise"), na.rm=TRUE), 
    		blueplus_percent = round(100 * blueplus_count / total_count, 1)) 
```

# What price did students in 431 pay for a haircut?

Let's summarize this graphically first. Perhaps a histogram?

```{r}
ggplot(data = day1, aes(x = haircut)) +
    geom_histogram()
```

## Improving the Histogram

Let's change the fill and color, and specify the size of the bins to be $10.

```{r}
ggplot(data = day1, aes(x = haircut)) +
    geom_histogram(binwidth = 10, 
                   fill = "blue", color = "white")
```

Can we improve the labels?

```{r}
ggplot(data = day1, aes(x = haircut)) +
    geom_histogram(binwidth = 10, 
                   fill = "blue", color = "white") +
    labs(x = "Price of Last Haircut in $", 
         y = "Number of Students",
    		 title = "Histogram of Prices for Most Recent Haircut",
    		 subtitle = "Students in 431: 2014-2018")
```

Can we eliminate that missing values issue?

```{r}
day1 %>%
    filter(!is.na(haircut)) %>%
ggplot(., aes(x = haircut)) +
    geom_histogram(binwidth = 10, 
    							 fill = "blue", color = "white") +
    labs(x = "Price of Last Haircut in $", 
    		 y = "Number of Students",
    		 title = "Histogram of Prices for Most Recent Haircut",
    		 subtitle = "Students in 431: 2014-2018")
```

## Numerical Summary of Haircut Prices

```{r}
day1 %>%
    select(haircut) %>%
    summary()
```

which can also be done with

```{r}
summary(day1$haircut)
```

The `mosaic` package has a useful `favstats` function...

```{r}
mosaic::favstats(day1$haircut)
```

But to get this in a pipeline, you'd need the `%$%` operator from the `magrittr` package...

```{r}
day1 %$%
    mosaic::favstats(haircut)
```

The `psych` package has a useful `describe` function...

```{r}
day1 %$%
    psych::describe(haircut)
```

The `Hmisc` package also has a useful `describe` function...

```{r}
day1 %$%
    Hmisc::describe(haircut)
```

## A more challenging question: Haircut Prices by Year

Can we build a separate plot for each year?

```{r}
day1 %>%
    filter(!is.na(haircut)) %>%
    ggplot(., aes(x = haircut, fill = year)) +
        geom_histogram(binwidth = 10, color = "white") +
        guides(fill = FALSE) +
        labs(x = "Price of Last Haircut in $", 
        		 y = "Number of Students") +
        facet_wrap(~ year)
```

We could use a comparison boxplot, but this requires us to get R to see `year` as a factor...

```{r}
ggplot(data = day1, aes(x = factor(year), y = haircut)) +
    geom_boxplot()
```

## Numerical Summary by Year?

```{r}
day1 %>% 
    filter(!is.na(haircut)) %>%
    group_by(year) %>%
    summarize(mean = mean(haircut), sd = sd(haircut), 
    					median = median(haircut))
```

# What is the relationship between sex and guess of my age?

```{r}
day1 %>%
    filter(!is.na(ageguess)) %>%
ggplot(., aes(x = sex, y = ageguess)) +
    geom_point() 
```

Let's restrict to 2018, color the points, and jitter them left-to-right a bit.

```{r}
day1 %>%
    filter(!is.na(ageguess), year == "2018") %>%
ggplot(., aes(x = sex, y = ageguess, col = sex)) +
    geom_jitter(width = 0.2) 
```

A boxplot would be the usual summary of a quantitative variable's association with a categorical variable. We'll add a title, too.

```{r}
day1 %>%
    filter(!is.na(ageguess), year == "2018") %>%
ggplot(., aes(x = sex, y = ageguess, fill = sex)) +
    geom_boxplot() +
    labs(title = "Guesses of Professor Love's Age, in 2018")
```

We might look at faceted histograms instead.

```{r}
day1 %>%
    filter(!is.na(ageguess), year == "2018") %>%
ggplot(., aes(x = ageguess, fill = sex)) +
    geom_histogram(bins = 10, col = "white") +
    facet_wrap(~ sex) + 
	  guides(fill = FALSE) +
    labs(title = "Guesses of Professor Love's Age, in 2018",
    		 subtitle = "His actual age was 51.5")
```

# What is the relationship between 431 students' pulse rate and hours of sleep?

Here, we're looking at two quantitative variables. A scatterplot is usually the best choice.

```{r}
ggplot(data = day1, aes(x = pulse, y = lastsleep)) +
    geom_point()
```

Let's look at the 2018 data only, and fit a model to predict last sleep on the basis of pulse rate.

```{r}
day1 %>%
    filter(year == "2018") %>%
    ggplot(., aes(x = pulse, y = lastsleep)) +
    geom_point() +
    geom_smooth(method = "loess") +
    labs(title = "Hours of Sleep as a function of Pulse Rate",
         subtitle = "with fitted loess smooth, students in the 2018 class")
```

We could instead restrict ourselves to a linear model for last sleep on the basis of pulse rate.

```{r}
day1 %>%
    filter(year == "2018") %>%
    ggplot(., aes(x = pulse, y = lastsleep)) +
    geom_point() +
    geom_smooth(method = "lm", col = "red") +
    labs(title = "Hours of Sleep as a function of Pulse Rate",
         subtitle = "with fitted linear model, students in 2018")
```

The correlation of `lastsleep` and `pulse` is likely to be of some interest. Note the use of both the `%>%` and `%$%` pipes in this case.

```{r}
day1 %>%
    filter(year == "2018") %$%
    cor(pulse, lastsleep)
```

## Does the linear model change much by year?

Here's the plot, color coding the models by year.

```{r}
day1 %>%
    filter(!is.na(pulse), !is.na(lastsleep)) %>%
    ggplot(., aes(x = pulse, y = lastsleep, 
    							color = factor(year),
    							group = factor(year))) +
    geom_point() + 
    geom_smooth(method = "lm", se = FALSE) +
	  theme_bw()
```

Here's the same plot of the models alone, and not showing the data (we just commented out the key line.)

```{r}
day1 %>%
    filter(!is.na(pulse), !is.na(lastsleep)) %>%
    ggplot(., aes(x = pulse, y = lastsleep, 
    							color = factor(year),
    							group = factor(year))) +
 #   geom_point() + 
    geom_smooth(method = "lm", se = FALSE) +
	  theme_bw()
```

Here's the same basic information, but faceted by year.

```{r}
day1 %>%
    filter(!is.na(pulse), !is.na(lastsleep)) %>%
    ggplot(., aes(x = pulse, y = lastsleep, 
    							group = factor(year))) +
    geom_point() + 
    geom_smooth(method = "lm", color = "red") +
    facet_wrap(~ year) +
	  theme_bw()
```