---
title: "Class 3 Live Analysis: Exploring Day 1 Survey"
author: "TEL"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    code_folding: show
---

```{r setup, message=FALSE}
knitr::opts_chunk$set(comment=NA)
options(width = 70)
```

```{r load_packages, message = FALSE, warning = FALSE}
library(magrittr); library(tidyverse)
## always need tidyverse, can include other packages too
```

# Load The Data

```{r}
day1 <- read.csv("surveyday1_2018.csv") %>% tbl_df()

day1
```

## Variables we're exploring today

```{r}
day1 %>%
  select(sex, favcolor, haircut, ageguess, pulse, lastsleep) %>%
  summary()
```

# What are the favorite colors of students in 431?

We can count all responses...

```{r}
day1 %>%
  count(favcolor)
```

We could restrict ourselves to 2018, and arrange them in descending order of the count...

```{r}
day1 %>%
  filter(year == 2018) %>%
  count(favcolor) %>%
  arrange(desc(n))
```

```{r}
day1 %>%
  select(favcolor) %>%
  table()
```

```{r}
table(day1$favcolor)
```

## A more challenging question

What is the percentage of students who preferred "blue" in each year?

```{r}
day1 %>%
  group_by(year) %>%
  summarize(blue_count = sum(favcolor == "blue", na.rm = TRUE),
            total_count = sum(!is.na(favcolor)),
            blue_percent = round(100 * blue_count / total_count, 1))
```

# What price did students pay for a haircut?

Let's try a histogram.

```{r}
ggplot(data = day1, aes(x = haircut)) +
  geom_histogram()
```

## Improving the Histogram

```{r}
ggplot(data = day1, aes(x = haircut)) +
  geom_histogram(binwidth = 10, fill = "navy", color = "white") +
  labs(x = "Price paid for last haircut in $",
       y = "# of students",
       title = "Histogram of Haircut Prices for 431 Students",
       subtitle = "2014-2018: Most Recent Haircut") +
  theme_bw()
```

