---
title: "431 Homework 8"
author: "Thomas E. Love"
date: "Due **2018-11-30** at noon. Version: `r Sys.time()`"
output:
    github_document:
        toc: yes
---

# General Instructions

This homework includes **six** questions. Be sure to respond to each question. 

# Question 1 (30 points)

Tell us about the most useful thing (an insight, example or key idea) that you learned from reading Nate Silver's *The Signal and the Noise*. Be sure to specify **in detail** how this insight/example/idea connects to your own life, or work. Specify at least one quote from the book that discuss this issue, and explain the context. Be sure that when we finish your essay we have a clear understanding of how reading this book has changed the way you think about the thing you describe or things related to it.

An appropriate response to this question will be 200-400 words long.

## A note on the specifications for the essay

In reading your essay, we will look for the following specifications to be met:

1. Length is between 200 and 400 words.
2. English is correctly used throughout, and there is no more than one typographical error and one grammatical / syntax error.
3. A key idea is identified and clearly stated that actually appears in *The Signal and the Noise*.
4. An accurate and properly cited quote from the book is provided that is relevant to the identified key idea.
5. The context for the quote within Silver's book is described in the student's essay.
6. The essay clearly specifies how the idea in the book has changed their way of thinking about something which is explained in the essay.
7. The essay is clearly written, in general.
8. The essay is interesting to read.

# General Description of the Data for Questions 2-6

Low dietary intake or low plasma concentrations of retinol, beta-carotene, or other carotenoids might be associated with increased risk of developing certain types of cancer. However, relatively few studies have investigated the determinants of plasma concentrations of these micronutrients. A cross-sectional study was designed to investigate the relationship between personal characteristics and dietary factors, and plasma concentrations of beta-carotene. Study subjects ($n = 300$) were patients who had an elective surgical procedure during a three-year period to biopsy or remove a lesion of the lung, colon, breast, skin, ovary or uterus that was found to be non-cancerous. 

Variable           | Description
-----------------: | -------------------------------------------------------
`subj_ID`  | Subject identification code (S-1001 to S-1300) 
`age` | Subject's age (in years)  
`sex` | Subject's sex (M = male, F = female) 
`smoking` | Smoking Status (Never, Former, or Current) 
`bmi` | Body-Mass Index ( weight in kilograms / [height in meters]^2^ ) 
`vitamin` | Vitamin Use (1 = Often, 2 = Sometimes, 3 = Never) 
`calories` | Number of Calories consumed (per day) 
`fat` | Number of grams of fat consumed (per day) 
`fiber` | Number of grams of fiber consumed (per day) 
`alcohol` | Number of alcoholic drinks consumed (per week) 
`cholesterol` | Number of milligrams of cholesterol consumed (per day) 
`betaplasma` | Plasma beta-carotene (in ng$/$ml) 

The `hw8_plasma` data set is available on our web site. It contains 300 observations. You will use a subset of 240 of those observations (selected at random) to fit your models, and the remaining 60 observations (those for which `holdout` = 1) to validate your model selection. 

In questions 2-6, you will build and evaluate a pair of multiple regression models for plasma beta carotene levels. To get started, I suggest this approach to loading the data.

```{r, message = FALSE}
library(tidyverse)

hw8_plasma <- read_csv("hw8_plasma.csv") %>%
    mutate_if(is.character, funs(as.factor(.))) %>%
    mutate(subj_ID = as.character(subj_ID))
```

Then use the following code to select a training sample from the data set (to be used in questions 2-5) and a test sample (to be used in question 6). Use `431008` as your seed as we have done here - that is what we'll do in the Answer Sketch.

```{r}
set.seed(431008)
hw8_training <- hw8_plasma %>% sample_n(240)
hw8_test <- anti_join(hw8_plasma, hw8_training, 
                      by = "subj_ID")
```

# Question 2 (15 points)

Use the `hw8_training` data frame to plot the distribution of the outcome of interest, which is `betaplasma`, and then plot the logarithm of `betaplasma`. Specify which of the two distributions better matches the desirable qualities of an outcome variable in a regression model. Whichever choice you make as to which outcome (`betaplasma` or `log(betaplasma)`) stick with it for the rest of this homework. 

# Question 3 (10 points)

Use the `hw8_training` data frame to build a model for your outcome (as decided in Question 2) using the following 4 predictors: `age`, `sex`, `bmi`, and `fiber`. Call that model `model_04`.

Summarize `model_04` and write a sentence or two to evaluate it. Be sure you describe the model's R^2^ value. Also, be sure to interpret the model's residual standard error, in context.

# Question 4 (10 points)

For your `model_04`, what is the estimated effect of being female, rather than male, on your outcome, holding everything else (age, bmi and fiber) constant. Provide and interpret a 95% confidence interval for that effect on your outcome.

# Question 5 (15 points)

Now use the `hw8_training` data frame to build a new model for your outcome (as decided in Question 2) using the following 10 predictors: `age`, `sex`, `smoking`, `bmi`, `vitamin`, `calories`, `fat`, `fiber`, `alcohol`, and `cholesterol`. Call that model `model_10`.

Compare `model_10` to `model_04` in terms of **adjusted** R^2^, and residual standard error. Which model performs better on these summaries, in the training sample?

# Question 6 (20 points)

Use the code provided in the [Project Study 2 Demonstration](https://github.com/THOMASELOVE/431-2018-project/tree/master/demo_study2) (section 14) to calculate and then compare the prediction errors made by the two models (`model_10` and `model_04`) you have generated. You should:

- Calculate the prediction errors in each case, then combine the results from the two models, following section 14.1 of the [Project Study 2 Demonstration](https://github.com/THOMASELOVE/431-2018-project/tree/master/demo_study2).
    - **HINT**: If you chose to transform the outcome variable back in Question 2, then you will need to estimate the predictions here back on the original scale of `betaplasma`, rather than on the logarithmic scale. That involves making predictions on the log scale, and then back-transforming them with the `exp` function before calculating the residuals and eventually the summary statistics.
- Visualize the prediction errors in each model, using the code in section of the Demo Project.
- Form the table comparing the model predictions, using the code in section 14.3. Compare the models in terms of MAPE, MSPE and maximum prediction error.

Based on your results, what conclusions do you draw about which model (`model_10` or `model_04`) is preferable? Is this the same conclusion you drew in Question 5?

